Verifying Spatial Properties of Stencil Computations using CamFort

### 2017-03-11 array17.tex
Array 17 first draft

\title{Verifying Spatial Properties of Stencil Computations}
%\titlebanner{DRAFT -- Do not distribute}
\authorinfo{}{}{}
%\preprintfooter{DO NOT DISTRIBUTE}

@@

\begin{abstract}
Stencil computations are a common programming pattern in numerical
  code and typically form the core of graphics and scientific
  computing applications. However, they are often complex and
  unwieldy, involving substantial fine-grained manipulation of array
  indices across multiple arrays and dimensions. This requires careful
  testing, which is almost always done manually and then discarded.
  In this paper, we propose an alternate approach:
  a specification language and automated tool for verifying the
  spatial behaviour of stencils. The specification language is based
  on the hypothesis that stencil computations read from arrays with a
  regular, fixed pattern which can be captured by a simple
  set of abstract combinators. These
  specifications can be embedded as annotations in source code,
  against which code can be checked for conformance. For legacy code,
  spatial specifications can be inferred, providing documentation and
  aiding future software maintenance. We evaluate our specification
  language and verification tool against
  a corpus of numerical Fortran code ($\sim$ 1 million lines) for
  which we generate $\sim$60,000 specifications, showing the vast majority
  of stencil computations indeed have a simple, regular, static shape. % We introduce the
  %language, checking and inference procedures, and its evaluation.
\end{abstract}

@@

In practice, the typical development procedure for complex
stencil computations involves some ad hoc testing to ensure that no
transcription mistakes have been made (\eg{}, by visual
inspections, or comparison against an manufactured or 
analytical solutions~\cite{farrell2010automated}). Typically such
testing is discarded once the code is seen to be correct.

@@

We provide a verification tool which can check the spatial correctness
of stencil computations against a specification. Furthermore, the tool
can automatically infer specifications and insert these as comments
into the source code. The main features and benefits of our approach
are as follows:
%
\begin{itemize}
%\item The main use case is that a programmer writes a specification
%  first (perhaps informed by their numerical analysis), then writes
%  the corresponding source code, and lastly has the tool check its
%  conformance to the specifications. This specify-and-check approach
%  removes the usually ad hoc testing effort, and provides the
%  opportunity for easy re-verification in the event of any refactoring
%  of software maintenance activities.
\item The primary use case is that a programmer writes a specification
  first before the stencil code. Our tool then checks conformance between
  the two. This specify-and-check approach reduces testing effort and
  future-proofs against bugs introduced during refactoring and maintenance.
%
%\item The specification provides a concise description which can be
%  preserved in the source code to provide documentation to future
%  developers, contrasting with ad hoc tests which are usually
%  discarded or specialised to the development environment. Separate from the
%  verification procedure, the infer-and-synthesise mode of the tool
%  provides a way to automatically generate additional useful
%  documentation for future users.

% Emphasis here is on documentation generation, so it makes little sense to
% compare this aspect of CamFort to ad hoc testing. Apples and oranges.
\item A specification concisely captures the access pattern of a
  stencil and turns it into documentation. The inference procedure
  efficiently produces specifications with no programmer intervention,
  automatically inserting specifications at appropriate places in the
  source code.
%
%\item Specifications are integrated into the code as comments for easy
%  integration with existing tools, as well as documentation systems (\eg{}, we
%  support Doxygen and FORD syntax).
%
\item Our specification format is deliberately abstract, with a small number of
  combinators that \emph{do not involve any indexing expressions}, \eg{}
  \texttt{a(i+1,j-1)}. This contrasts with other specification
  approaches, \eg{}, deductive verification tools such as ACSL, where
  specifications must also be in terms of array-indexing expressions. Therefore, any low-level mistakes
  that could be made whilst programming complex indexing code could also be
  made when writing its specification. Our specifications are more
  abstract and lightweight
  with the aim of aiding adoption by scientists and preventing indexing errors.
\end{itemize}

@@

In this paper, we make the following contributions:
%
\begin{itemize}
\item We introduce a specification language for stencils
  that captures common forms of data access patterns
  (\cref{sec:lang});

\item We detail checking and inference algorithms for our
  specifications (\cref{sec:analysis}), derived from a
  denotational model for the specification language
  (\cref{sec:semantics});

\item We provide an implementation of our approach as an extension to
  CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;

\item We report on a quantitative study of stencil computations on a
  corpus of numerical Fortran program
  (\cref{sec:evaluation}), totalling one million lines of code. Our tool
  identifies and infers specifications for roughly 60,000 stencil computations in the
  corpus. Approximately 5,000 of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors. This validates
  our hypothesis that the majority of stencil computations have a regular
  shape, and validates the design of our language in its capability to capture
  many core patterns.

\item We give a verification case study of a
  stencil computation commonly used in scientific computing,
  simulating programming errors (\Cref{sec:jacobi}). Our approach detects all
  possible simulated programming errors for this problem.
\end{itemize}
%

@@

We give various definitions used in the paper then
\cref{subsec:syntax} outlines the syntax of stencil specifications
with an informal introduction. \cref{sec:semantics}
gives a semantic model via a set-based interpretation which makes
concrete the meaning of specifications. \cref{sec:eqs} then provides
a syntactic equational theory for specifications which is
sound with respect to the model.
%sub-specification for which the
%equality and sub-specification relation are proven sound.

@@

\begin{definition}[Stencil computations]
\label{def:stencil}
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
  collection of arrays of arbitrary (possibly differing)
  dimensionalities.
  A \emph{stencil computation} comprises an iteration by a set $I$ of at
  most $n$ induction variables
  over a subset of the index space of $a$. Each iteration
  determines the elements of $a$ by an
  assignment $a(\bar{e}) = e_r$ where each $e \in
  \bar{e}$ is either a neighbourhood or constant index.
  For all $b \in \bar{b}$ and array subscripts $b(\bar{e}')$
  that flow to $e_r$, each $e' \in \bar{e}'$ is
  either a neighbourhood or absolute index.

  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
\end{definition}
%
% specifying the shape of indexing expressions for
%arrays $\bar{v}$ that are simultaneously permitted and expected within
%the code to which the specification is given.
%
%\marginpar{\dnote{Is it worth having this informal consistency here?}}

For now we informally describe what it means for a stencil kernel to
be consistent with a specification. This will be formalised in
\Cref{sec:semantics} and algorithmically described in
\Cref{sec:analysis}.  Specifications are associated to a stencil
kernel and are declared for one or more array variables $\bar{v}$
within that kernel.  A stencil specification defines a \emph{region}
associated to an array variable $a$ against which a stencil kernel
should be \emph{consistent}. The region defined by a specification is
an $n$-dimensional rectilinear shape in a discrete space crossing the
origin (which represents a collection of indices which are just 
induction variables). Consistency is then a
two way requirement. All array subscripts on $a$ contributing to the
kernel should be consistent with the region. Conversely, all parts of
the region should have an index corresponding it.

@@

A \term{forward} region of depth $n$ specifies a contiguous
region in dimension $d$ starting at the origin. This corresponds
to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
for some induction variable $i$. Similarly, a
\term{backward} region of depth $n$ corresponds to contiguous indices
from $i$ to $i - n$ and \term{centered}
of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
stencil specifies a neighbour index $i$. For example, the
following shows four specifications with four consistent stencil
kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
!= stencil backward(depth=2, dim=1) :: b
e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
!= stencil centered(depth=1, dim=2) :: c
e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
!= stencil pointed(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
Not every dimension needs to be specified, \eg{},
specifications on lines $1$, $3$, and $7$ leave some dimensions unspecified.

@@

The region operators \term{+} and \term{*} respectively combine
regions disjunctively and conjunctively.

The conjunction of two regions $r \term{*} s$ means that any indices
in the specified code must be consistent with both $r$ and $s$
simultaneously. Dually, for every pair of components in the regions
$r$ and $s$ there must be an index consistent with both.
For example, the following \emph{nine-point stencil}
has a specification given by the product of two \texttt{centered}
regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%%
This pattern is common in image convolution applications.
The specification ranges over the
values that flow to the array subscript on the left-hand side,
and so ranges over the intermediate assignments to \term{x},
\term{y}, and \term{z}. Each index in the code is consistent
with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
is within the centered region in dimension $1$ and the centered region
in dimension $2$.

@@

The disjunction of two regions $r \term{+} s$ means that any indices
in the specified code must be consistent with either of $r$ or $s$.
Dually, every part of region $r$ must have a
corresponding index and similarly for $s$ independently. For
example, the following gives the specification of a five-point
stencil which is the sum of two compound \texttt{pointed} and
\texttt{centered} regions in each dimension:
%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1)*pointed(dim=2)    + centered(depth=1, dim=2)*pointed(dim=1) :: a
b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
                   + a(i,j-1) + a(i,j+1)
\end{minted}
%%
Here the left-hand side of \texttt{+} says that when the second dimension
(induction variable $j$) is fixed at the origin, the first dimension
(induction variable $i$) accesses the immediate vicinity of the origin
(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
This reflects the symmetry under rotation of the five-point stencil.

@@

\section{A model for specifications}
\label{sec:semantics}

We define a simple set-theoretic denotational model of the semantics
of our specification language. This has a number of purposes:
(1) it serves to explain the meaning of our specifications
(2) it is used in the inference and checking algorithms
(\Cref{sec:analysis}); (3) justifies an equational theory for
specifications in the next section; and (4) can be used to guide
correct implementations. The model domain is over sets of vectors
which we call \emph{index schemes}.

\begin{definition}
  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
  (\ie{}, the constant integer $c$ in an indexing expression of the
  form $i \pm c$, \Cref{def:neighbour}) and $\infty$ is a wild-card
  representing both constant indices and unspecified indices.
  For an index scheme $u$, the $i^{th}$ element is denoted $u_i$. We
  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
\end{definition}

Index schemes abstract array subscripts.
Our model consists of index schemes explaining the range of possible
indexing behaviours which are consistent with the specification.
The relationship between source language
and index schemes
 is defined by \textsf{schematic} function,
 which forms part of checking and inference (\Cref{sec:analysis}).

\newcommand{\schemeOne}{\textsf{s}_I}
%\newcommand{\schemeOne}{\textsf{schema$^1_I$}}
\newcommand{\schemeN}{\textsf{schematic}_I}
%\newcommand{\schemeN}{\textsf{schema$^n_I$}}
\begin{definition}%[Schematic]
\label{def:schematic}
Let $\schemeOne$ be a partial function, parameterised by a set of
induction variables $I$, mapping source language expressions to
$\domainVal$ values:
%%
\begin{align*}
\schemeOne(e)
 & =  \begin{cases}
c & e \equiv i + c \;\; \wedge \;\; i \in I \\
-c & e \equiv i - c \;\; \wedge \;\; i \in I \\
\infty & \textit{$I \cap \textit{FreeVariables}(e) = \emptyset $} %\\
%\bot   & \textit{otherwise}
\end{cases}
\end{align*}
In the
first two cases, expressions that are a constant offset of an
induction variable (\ie{}, neighbourhood indices) are mapped to the
offset $c$ or $-c$ $\in \domainVal$. In the third case, if the
expression is not defined in terms of an induction variable, it
is mapped to $\infty \in \domainVal{}$. Otherwise, $\schemeOne(e)$
is undefined ($\bot$).

The partial function $\schemeN$, also parameterised by a set of
induction variables $I$, uses $\schemeOne$ to
 map from the syntax of indices to index schemes
$\in (\domainVal)^n$. It is defined on the syntax of indices $\bar{e} = (e_1, ..., e_n)$ as:
\begin{align*}
\schemeN(e_1, .., e_n) & =
[\schemeOne(e_1) ... \schemeOne(e_n) ]
\; \textit{iff} \; \forall x . \schemeOne(e_x) \neq \bot
\end{align*}
%
Thus, if each expression in an index is a neighbourhood or constant
index, $\schemeN$ maps it to an index scheme.
\end{definition}

\begin{definition}[Generalisation]
\label{def:generalisation}
Given two schemes $v, w$ of dimensionality $n$,
then $w$ is said to \emph{generalise} $v$
(or conversely $v$ \emph{specialises} $w$),
written $v \preceq_n w$ defined:
\[
v \preceq_n w \;\; \Leftrightarrow  \;\;
  \forall i\!\in\!\{ 1, \ldots, n \} . \, (w_i = \infty \, \vee \, v_i = w_i)
\]
%%
where $\preceq_n$ is a reflexive, transitive relation on
index schemes $\domainVal^n$. Thus,
 $\infty$ generalises any index in the same dimension.
%
\end{definition}

\begin{example}
Consider the following array subscript to index schemes
conversions:
%
\begin{align*}
\begin{array}{ll}
\schemeN{(i, j+1)} & =\vtwoh{0}{1} \\
\schemeN{(1, j+1)} & = \vtwoh{\infty}{1} \\
\schemeN{(i*2, j)} & = \bot
\end{array}
\end{align*}
There is one generalisation between these
schemes, namely that $\vtwoh{0}{1} \preceq_2 \vtwoh{\infty}{1}$.
The last example is undefined as its first expression is a
non-neighbour relative index.
\end{example}

%Our terminology of \emph{index schemes} and
%\emph{generalisation} is by analogy
%with the \emph{type schemes} and \emph{type generalisation} of
%polymorphic types in ML~\cite{milner1978theory}. Our notion is
%much more rigid however, as index schemes do not bind names.

\paragraph{Shorthand notation}

We use abbreviations
 \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
for \term{forward}, \term{backward},
\term{centered}, and \irrefl{} respectively. Furthermore, 
we quantify over the absence or presence of the \irrefl{}
attribute on region constants as a boolean. For example,
\term{fwd} is syntactic sugar defined:
%%
\begin{align*}
\hspace{-0.5em}
& \stenFwdSR{n}{d}{p}
\; := \; \\[-0.4em]
& \quad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, p \\[-0.1em]
\stenFwd{n}{d}{}  & \textit{if} \;\,  p
\end{cases}
\end{align*}
%
$\textit{or} \;\; \stenFwdS{n}{d} := \stenFwd{n}{d}{}$
and similarly for \term{bwd} and \term{cen}.

\subsection{Denotational model}

An interpretation function $\interp{-}_n$ provides the model 
(where $n$ is the maximum dimensionality of the specification being
modelled) mapping closed\footnote{That is, we assume there are no
  occurrences of \textit{rvar} in a specification being modelled.  Any
  \emph{open} specification containing region variables can be made
  closed by straightforward syntactic substitution with a (closed)
  \textit{region}.} specifications to sets of
$n$-dimensional index schemes. The interpretation is overloaded on
\emph{regions} in \Cref{fig:region-model} and on the top-level
of a specification \textit{spec} in \Cref{fig:spatial-model}.
Various intermediate notions are used.

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.
Thus, a single-entry vector is an index scheme generalising 
all other schemes that have $1$ as their $r^{th}$ entry,
describing a neighbour offset of $1$.
\end{definition}

\begin{definition}
A \emph{zeroed single-entry vector} of size $n$ denoted
$\zeroEntry{r}{n}$ is a vector where the $r^{th}$ entry is $0$ and all others
are $\infty$ \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
Akin to single-entry vectors, zeroed single-entry vectors
generalise all other schemes
that have $0$ as their $r^{th}$ entry, representing indices at
the ``origin'' in $r$.
\end{definition}

\begin{figure}[t]
\vspace{-0.9em}
\input{region-model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Model of regions,
$\interp{-}_n : \textit{region} \rightarrow \mathcal{P}(\domainVal{}^n)$}
\label{fig:region-model}
\vspace{-1em}
\end{figure}

\noindent
The first four equations of \Cref{fig:region-model}
model region constants as sets of indexing
schemes. We illustrate with an example.

\begin{example}
For a $2$-dimensional stencil computation, then
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\[-0.2em]
& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
\end{align*}
Note, $\infty$ is absorbing for multiplication above \eg{} $2
 \infty = \infty$.
\end{example}
\noindent
The final two equations of \Cref{fig:region-model} give the models
of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
 union of the models of subterms. The model of
$\term{*}$ takes a ``tensor'' $\otimes$ of subterm models. The tensor is
more involved. We give some informal intuition first.

% r+s means indices must be consistent with either r or s
%           and there are indices why are generalised by r and s
% (there are indices which are consistent with every part of r or s)
% (every part of r and s independently have an index which is generalised by it)

% r*s means indices must be consistent simultaneously with both r and s
%           and there are indices which are generalised
%

\begin{figure}
\vspace{-1em}
\input{spatial-model.tex}
\caption{%\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\vspace{-1em}
\end{figure}

\paragraph{$\otimes$ of models}

The intuition behind $\otimes$ is that it takes all possible pairs of
index schemes, and treats each pair as lower and upper bounds of an
$n$-dimensional rectangle. From these bounds,
the remaining vertices of the $n$-dimensional rectangle are
generated (like a bounding box) \eg{}:
%%
\begin{align*}
{\small{
\{\vtwohs{1}{2},\vtwohs{5}{6}\,\}\!\otimes\!\{\vtwohs{3}{4}\}\!=\!\{\vtwohs{1}{2},\vtwohs{3}{2},\vtwohs{1}{4},\vtwohs{3}{4},
   \vtwohs{3}{6},\vtwohs{5}{4},\vtwohs{5}{6}\,\}
}}
\end{align*}
%%
The tensor takes a Cartesian product of the two models,
giving all pairs of schemes which are combined using the
\emph{pairwise permutation} written $\bowtie$ (defined later).
Pairwise permutation generates all possible interleavings of two
vectors. Some care is then taken over how this interacts with $\infty$.
In the presence of $\infty$, generate all interleavings as if $\infty$ is
an integer, then filter out vectors with $\infty$ at any coordinate, \eg{}
\begin{align*}
\{\vtwoh{1}{\infty}\} \otimes \{\vtwoh{3}{4}\} =
\{\vtwoh{1}{4}, \vtwoh{3}{4}\}
\end{align*}
%
where $\vtwoh{1}{\infty}$ and $\vtwoh{3}{\infty}$ have been filtered.

\begin{definition}%[Constrained dimensions]
For an $n$-dimensional model $M$, its \emph{constrained dimensions}
are the set of dimension identifiers where at least one scheme $u \in M$ has
a non-$\infty$ value in that dimension:
%
\begin{align*}
\effdims{M}{n} =
\bigcup_{u \in M} \{i \mid i \in \{1,\ldots,n\} \, \wedge \, u_i \neq
  \infty\}
\end{align*}
For example, $\effdims{\{\vtwoh{0}{\infty},\vtwoh{\infty}{\infty}\}}{2}
 = \{1\}$.
\end{definition}
\begin{definition}The tensor $\otimes_n$ of $n$-dimensional
models is:
\begin{alignat*}{2}
& M \otimes_n N =
  \Bigg\{x \; \Bigg| \;
    \parbox{5.2cm}{$(u, v) \in M \times N,$ \\[0.1em]
                  $\;\; i \in \{1, \ldots, 2^n\}, \,
                  x = (u \bowtie v)_i,$ \\[0.1em]
                  $\;\; j \in (\effdims{M}{n} \cup
                   \effdims{N}{n}),$\\[0.1em]
                   $\;\; x_j \neq \infty$
                  } \Bigg\}
\end{alignat*}
The second guard binds $j$ to the constrained
dimensions for $M$ unioned with the constrained dimensions for $N$.
That is, every $j$ corresponds to a dimension in which
either $M$ or $N$ has a scheme $u$ where $u_j \neq \infty$,
and thus is constrained.
The \emph{pairwise permutation} $u \bowtie v$
builds a $2^n \times n$ matrix
of all possible $n$-vectors generated by
non-deterministically picking for each $j^{th}$ entry either
$u_j$ or $v_j$ \eg{}
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\left[
\setlength{\arraycolsep}{0.42em}
\begin{array}{cccccccc}
0 & 0 & 0 & 0 & 3 & 3 & 3 & 3 \\
1 & 1 & 4 & 4 & 1 & 1 & 4 & 4 \\
2 & 5 & 2 & 5 & 2 & 5 & 2 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
correspond to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. The pairwise permutation
matrix is thus defined in terms of the logical
matrix $b$ where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$ as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
The permuted schemes $x = (u \bowtie v)_i$ added to the model
$M \otimes N$ are only those without $\infty$ in
the constrained dimensions given by each model ($x_j \neq
\infty$). This eliminates all permutations with
$\infty$ in a constrained dimension position.
\end{definition}

\paragraph{Model of spec}

Regions are modelled as sets of index
schemes, that is
$\interp{\textit{region}}_n \in \mathcal{P}((\domainVal{})^n)$.
\Cref{fig:spatial-model} defines the top-level model of the
\textit{spec} syntax, which includes the multiplicity and
approximation modifiers. The domain of this interpretation
is $\textsf{Mult}(\textsf{Approx}(\mathcal{P}((\domainVal{})^n)))$
where $\textsf{Mult}$ and $\textsf{Approx}$ are labelled variants
labelling our set-based model with additional information.

\begin{definition} \textsf{Mult}
and \textsf{Approx} are parametric labelled variant types
with injections given by their definition:
%
\begin{align*}
\textsf{Mult} \, a & = \textsf{mult} \, a \,\mid\, \textsf{only} \, a \\
\textsf{Approx} \, a & = \textsf{exact} \, a \,\mid\, \textsf{lower} \, a \,\mid\,
\textsf{upper} \, a \,\mid\, \textsf{both} \, a \, a
\end{align*}
\eg{}, $\mathsf{lower}$ is an injection
$\mathsf{lower} : a \rightarrow \mathsf{Approx} \, a$ etc. The
$\textsf{Mult}$ type corresponds to the presence or absence of the
\term{readOnce} modifier as shown in \Cref{fig:spatial-model}.
The $\textsf{Approx}$ type corresponds to the presence or absence of
the spatial approximation modifier, with \textsf{exact} when there is
no such modifier and \textsf{lower} and \textsf{upper} for
\term{atLeast} and \term{atMost}. Later, it will be useful to model
pairs of lower and upper bounds provided by \textsf{both} but which is
not used
here.% These are used for labelled models, and other components later in.
\label{def:mult-and-approx}
\end{definition}

@@

Specifications have a theory of equality, $\equiv$, providing axiomatic
semantics for the language. This enables users to rewrite their specification
and provides a way to automatically simplify
specifications after inference (\Cref{subsec:inference}).
Additionally, specifications have a theory of approximation,
$<:$, that can be used by the users to weaken specifications for succinctness in
the source.

>>>>

\title{Verifying Spatial Properties of Stencil Computations using CamFort}
%\titlebanner{DRAFT -- Do not distribute}
\authorinfo{}{}{}
%\preprintfooter{DO NOT DISTRIBUTE}

@@

\begin{abstract}
  Stencil computations are a common programming pattern in numerical code and
  typically form the core of graphics and scientific computing applications.
  However, they are often complex and unwieldy, involving substantial
  fine-grained manipulation of array indices across multiple arrays and
  dimensions. This requires careful testing, which is almost always done
  manually and then discarded. In this paper, we propose an alternate approach:
  a specification language and automated tool for verifying the spatial
  behaviour of stencils. The specification language is based on the hypothesis
  that stencil computations read from arrays with a regular, fixed pattern
  which can be captured by a simple set of abstract combinators. These
  specifications can be embedded as annotations in source code, against which
  code can be checked for conformance. For legacy code, spatial specifications
  can be inferred, providing documentation and aiding future software
  maintenance.

  We evaluate our specification language and verification tool
  against a corpus of numerical Fortran code ($\sim$ 1 million lines) for which
  we generate $\sim$ 60,000 specifications, showing the vast majority of stencil
  computations indeed have a simple, regular, static shape.
\end{abstract}

@@

In practice, the development process for complex stencil computations involves
ad hoc testing to ensure that no transcription mistakes have been made (\eg{},
by visual inspections, or comparison against an manufactured or analytical
solutions~\cite{farrell2010automated}). Typically such testing is discarded
once the code is seen to be correct.

@@

We provide a verification tool which can check the spatial correctness of
stencil computations against a specification. Furthermore, the tool can
automatically infer specifications and insert these as comments into the source
code. The main features and benefits of our approach are as follows:
%
\begin{itemize}
\item The primary use case is that a programmer writes a specification first
  before the stencil code. Our tool then checks conformance between the two.
  This specify-and-check approach reduces testing effort and future-proofs
  against bugs introduced during refactoring and maintenance.
%
\item A specification concisely captures the access pattern of a stencil and
  turns it into documentation. The inference procedure efficiently produces
  specifications with no programmer intervention, automatically inserting
  specifications at appropriate places in the source code.
%
\item Our specification format is deliberately abstract, with a small number of
  combinators that \emph{do not involve any indexing expressions}, \eg{}
  \texttt{a(i+1,j-1)}. This contrasts with other specification approaches,
  \eg{}, deductive verification tools such as ACSL, where specifications must
  also be in terms of array-indexing expressions. Therefore, any low-level
  mistakes that could be made whilst programming complex indexing code could
  also be made when writing its specification. Our specifications are more
  abstract and lightweight with the aim of aiding adoption by scientists and
  preventing indexing errors.
\end{itemize}

@@

In this paper, we solely focus on the definition and the usage of the language
and the acoompanying verification tool and omit the underlying mathematical
model and implementation details due to space considerations. We make the
following contributions:
%
\begin{itemize}
\item We introduce a specification language for stencils that captures common
  forms of data access patterns (\cref{sec:lang});
%
\item We provide an implementation of our approach as an extension to
  CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;
%
\item We report on a quantitative study of stencil computations on a corpus of
  numerical Fortran program (\cref{sec:evaluation}), totalling one million
  lines of code. Our tool identifies and infers specifications for roughly
  60,000 stencil computations in the corpus. Approximately 5,000 of the
  stencils we found are non-trivial, corresponding to code which is a possible
  source of errors.
%
\item We give a verification case study of a stencil computation commonly used
  in scientific computing, simulating programming errors (\Cref{sec:jacobi}).
  Our approach detects all possible simulated programming errors for this
  problem.
\end{itemize}
%

@@

We give various definitions used in the paper then \cref{subsec:syntax}
outlines the syntax of stencil specifications. \cref{sec:eqs} then provides a
syntactic equational theory for specifications which is sound with respect to
the underlying model.

@@

\texttt{BEGIN: NEEDS REWRITING}
\begin{definition}[Stencil computations]
\label{def:stencil}
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a collection of
  arrays of arbitrary (possibly differing) dimensionalities.  A \emph{stencil
  computation} comprises an iteration by a set $I$ of at most $n$ induction
  variables over a subset of the index space of $a$. Each iteration determines
  the elements of $a$ by an assignment $a(\bar{e}) = e_r$ where each $e \in
  \bar{e}$ is either a neighbourhood or constant index.  For all $b \in
  \bar{b}$ and array subscripts $b(\bar{e}')$ that flow to $e_r$, each $e' \in
  \bar{e}'$ is either a neighbourhood or absolute index.

  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
\end{definition}

We informally describe what it means for a stencil kernel to be consistent with
a specification. This will be algorithmically described in \Cref{sec:analysis}.
Specifications are associated to a stencil kernel and are declared for one or
more array variables $\bar{v}$ within that kernel.  A stencil specification
defines a \emph{region} associated to an array variable $a$ against which a
stencil kernel should be \emph{consistent}. The region defined by a
specification is an $n$-dimensional rectilinear shape in a discrete space
crossing the origin (which represents a collection of indices which are just
induction variables). Consistency is then a two way requirement. All array
subscripts on $a$ contributing to the kernel should be consistent with the
region. Conversely, all parts of the region should have an index corresponding
it.

\texttt{END: NEEDS REWRITING}

@@

A \term{forward} region of depth $n$ specifies a contiguous region in dimension
$d$ starting at the origin. This corresponds to specifying neighbourhood
indices in dimension $d$ ranging from $i$ to $i + n$ for some induction
variable $i$. Similarly, a \term{backward} region of depth $n$ corresponds to
contiguous indices from $i$ to $i - n$ and \term{centered} of depth $n$ from $i
- n$ to $i + n$. A \term{pointed} stencil specifies a neighbour index $i$. For
example, the following shows four specifications with four consistent stencil
kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
%%
\begin{minted}{fortran}
!= stencil backward(depth=2,dim=1) :: a
e(i, 0) = a(i, 0) + a(i-1, 0) + a(i-2, 0)
!= stencil forward(depth=2,dim=1)*pointed(dim=2) :: b
e(i, j) = b(i, j) + b(i+1, j) + b(i+2, j)
!= stencil centered(depth=1,dim=2) :: c
e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
!= stencil pointed(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
Every dimension with neighbourhood indices need to be specified. The remaining
non-conforming indices are not included in the specification, \eg{},
specifications on lines $1$ and $7$ leave some dimensions unspecified since they
do not conform to definition neighbourhood indices.

@@

The region operators \term{+} and \term{*} respectively combine regions
permissively and restrictively.

The restrictive combination of two regions, $r \term{*} s$, means that any
indices in the specified code must be consistent with both $r$ and $s$
simultaneously. Dually, for every pair of components in the regions $r$ and $s$
there must be an index consistent with both. For example, the following
\emph{nine-point stencil} has a specification given by the product of two
\texttt{centered} regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%%
This pattern is common in image convolution applications. The specification
ranges over the values that flow to the array subscript on the left-hand side,
and so ranges over the intermediate assignments to \term{x}, \term{y}, and
\term{z}. Each index in the code is consistent with both specifications
simultaneously, \eg{}, \texttt{a(i-1, j+1)} is within the centered region in
dimension $1$ and the centered region in dimension $2$.

@@

The permissive combination of two regions $r \term{+} s$ means that any indices
in the specified code must be consistent with either of $r$ or $s$. Dually,
every part of region $r$ must have a corresponding index and similarly for $s$
independently. For example, the following gives the specification of a
five-point stencil which is the sum of two compound \texttt{pointed} and
\texttt{centered} regions in each dimension:
%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1)*pointed(dim=2)    + centered(depth=1, dim=2)*pointed(dim=1) :: a
b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
                   + a(i,j-1) + a(i,j+1)
\end{minted}
%%
Here the left-hand side of \texttt{+} says that when the second dimension
(induction variable $j$) is fixed at the origin, the first dimension (induction
variable $i$) accesses the immediate vicinity of the origin (to depth of one).
The right hand side of \texttt{+} is similar but the dimensions are reversed.
This reflects the symmetry under rotation of the five-point stencil. As
presented in \cref{sec:evaluation}, this symmetry is common to stencil
computations used in practice, which makes it easier to write specifications.

@@

@@

Specifications have a theory of equality, $\equiv$, providing axiomatic
semantics for the language. This enables users to rewrite their specification
and provides a way to automatically simplify specifications after inference.
Here we only focus on the former and omit mathematical implications of theory.
Additionally, specifications have a theory of approximation, $<:$, that can be
used by the users to weaken specifications for succinctness in the source.
